Indicators for Ecosystem Condition in Norway

### 2023-06-15 DRAFT_ndvi_wetland.Rmd
markdown until data handling

<!-- Load all you dependencies here -->
```{r setup, include=FALSE}
library(sf)
library(plyr)
library(stringr, lib.loc = "/usr/lib/R/site-library")
library(tidyverse, lib.loc = "/usr/local/lib/R/site-library")
library(data.table)
library(lubridate)
library(RColorBrewer)
library("gridExtra") 
library(ggridges)
library(tmap)
library(tmaptools)
library(raster)
library(stars)
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
```

@@

In this draft development for a wetland NDVI indicator for ecological condition we approach these challenges in the following ways:
1) We use ecosystem occurrence data from NiN-mapping to guide the extraction of satellite NDVI data for pixels representing the right ecosystem. However, the ecosystem occurrences in the NiN data are highly spatially biased and thus cannot provide a good and unbiased spatial representation for the assessment area. Our work therefore represents a concept and method development which may be applied in an assessment of ecological condition once spatially unbiased occurrence data or a national map of wetland ecosystems are available.
2) We include Sentinel-II imagery for wetland ecosystems as the resolution for the bands underlying the NDVI product here is 10 x 10 m which is smaller than almost all wetland polygons in the NiN-data. However, since Sentinel-II only goes back to 2015, we lose much of the time series aspect as compared to the MODIS data. In addition, we still explore NDVI data from the MODIS and Landsat satellites (see more details below).

@@

- NiN ecosystem data from 'Naturtypekartlegging etter Miljødirektoratets innstruks' to guide extraction of NDVI data for the right locations and to provide information on the field-assessed ecological condition of mapped polygons
- Sentinel-II satellite NDVI data
- MODIS satellite NDVI data
- Landsat satellite NDVI data

@@

(2) Sentinel-II satellite NDVI data:

@@

The NDVI data from either satellite follow the spatial distribution and bias status of the NiN-data. Due to the larger number of excluded polygons in the NDVI data from MODIS (because of the large pixel size in MODIS, see above), the MODIS NDVI data for the wetland occurrences in the NiN data likely are even more spatially biased than it is the case for the NDVI data from Sentinel-II and Landsat.

@@

Sentinel-II satellite NDVI data span the years 2015-2022.
MODIS satellite NDVI data span the years ????-????.
Landsat satellite NDVI data span the years ????-????.

@@

## Analyses
### Data sets

#### NiN data
We read the NiN data for this document from a cache at the NINA server...
```{r, echo = F}
# Add data from cache
nin<-readRDS(paste0(here::here(),"/data/cache/nin.RDS"))
```

...but it can also be directly downloaded from Miiljødirektoratets kartkatalog (https://kartkatalog.miljodirektoratet.no/Dataset)
```{r, echo = T, eval = F}

# st_layers(dsn = "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/Naturovervaking_eksport.gdb")
# ANO.sp <- st_read("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/Naturovervaking_eksport.gdb",
#                    layer="ANO_Art")
# ANO.geo <- st_read("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/Naturovervaking_eksport.gdb",
#                   layer="ANO_SurveyPoint")
# head(ANO.sp)
# head(ANO.geo)
```

>>>>

<!-- Load all you dependencies here -->
```{r setup, include=FALSE}
library(downloader)
library(data.table)
library(lubridate)
library(sf)
library(plyr)
library(stringr, lib.loc = "/usr/lib/R/site-library")
library(tidyverse, lib.loc = "/usr/local/lib/R/site-library")
library(RColorBrewer)
library("gridExtra") 
library(ggridges)
library(tmap)
library(tmaptools)
library(betareg)
library(StepBeta)
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
```

@@

In this draft development for a wetland NDVI indicator for ecological condition we approach these challenges in the following ways:
1) We use ecosystem occurrence data from NiN-mapping to guide the extraction of satellite NDVI data for pixels representing the right ecosystem. However, the ecosystem occurrences in the NiN data are highly spatially biased and thus cannot provide a good and unbiased spatial representation for the assessment area. Our work therefore represents a concept and method development which may be applied in an assessment of ecological condition once spatially unbiased occurrence data or a national map of wetland ecosystems are available.
2) We include Sentinel-2 imagery for wetland ecosystems as the resolution for the bands underlying the NDVI product here is 10 x 10 m which is smaller than almost all wetland polygons in the NiN-data. However, since Sentinel-2 only goes back to 2015, we lose much of the time series aspect as compared to the MODIS data. In addition, we still explore NDVI data from the MODIS and Landsat satellites (see more details below).

@@

- NiN ecosystem data from 'Naturtypekartlegging etter Miljødirektoratets innstruks' to guide extraction of NDVI data for the right locations and to provide information on the field-assessed ecological condition of mapped polygons
- Sentinel-2 satellite NDVI data
- MODIS satellite NDVI data
- Landsat satellite NDVI data

@@

(2) Sentinel-2 satellite NDVI data:

@@

The NDVI data from either satellite follow the spatial distribution and bias status of the NiN-data. Due to the larger number of excluded polygons in the NDVI data from MODIS (because of the large pixel size in MODIS, see above), the MODIS NDVI data for the wetland occurrences in the NiN data likely are even more spatially biased than it is the case for the NDVI data from Sentinel-2 and Landsat.

@@

Sentinel-2 satellite NDVI data span the years 2015-2022.
MODIS satellite NDVI data span the years ????-????.
Landsat satellite NDVI data span the years ????-????.

@@

## Analyses
### Data sets
The analyses in this document make use of the following data sets:
-NiN data (Naturtyper etter Miljødirektoratets innstruks)
-Sentinel-2 NDVI data
-MODIS NDVI data
-Landsat NDVI data

#### NiN data and geometry for Norway and the five regions
We read the NiN data for this document from a cache at the NINA server...
```{r, echo = F}
# Add NiN data from cache
nin <- readRDS(paste0(here::here(),"/data/cache/nin.RDS"))
```

...but it can also be directly downloaded from Miljødirektoratets kartkatalog (https://kartkatalog.miljodirektoratet.no/Dataset)
```{r, echo = T, eval = F}

url <- "https://nedlasting.miljodirektoratet.no/Miljodata//Naturtyper_nin/FILEGDB/4326/Naturtyper_nin_0000_norge_4326_FILEGDB.zip"
download(url, dest="P:/41201785_okologisk_tilstand_2022_2023/data/Naturtyper_nin_0000_norge_4326_FILEGDB.zip", mode="wb") 
unzip ("P:/41201785_okologisk_tilstand_2022_2023/data/Naturtyper_nin_0000_norge_4326_FILEGDB.zip", 
       exdir = "P:/41201785_okologisk_tilstand_2022_2023/data/nin_data")
st_layers("P:/41201785_okologisk_tilstand_2022_2023/data/nin_data/Naturtyper_nin_0000_norge_4326_FILEGDB.gdb")
nin <- st_read("P:/41201785_okologisk_tilstand_2022_2023/data/nin_data/Naturtyper_nin_0000_norge_4326_FILEGDB.gdb",
                 layer="naturtyper_nin_omr")
```
And we upload maps for Norway and the five regions: Southern, Western, Eastern, Central, adn Northern Norway
```{r}
# Import region- og Norgeskart, make sure they have the same geometry as the nin data
nor <- st_read("data/outlineOfNorway_EPSG25833.shp")%>%
  st_as_sf() %>%
  st_transform(crs = st_crs(nin))

reg <- st_read("data/regions.shp")%>%
  st_as_sf() %>%
  st_transform(crs = st_crs(nin))

# combine the Norway and the region maps
regnor <- st_intersection(reg,nor)
```

#### NDVI data
All NDVI data are calculated mean NDVI values for each respective NiN polygon from each available Sentinel-2, MODIS, or Landsat image in Google Earth Engine (GEE).

##### Sentinel-2 NDVI data
The GEE code can be seen here: https://code.earthengine.google.com/2ceb0c3e03adade9e6f6d0903184b8c4
The image collection contains Sentinel imagery from June, July and August 2015-2022
To not exceed GEE memory limits, the exported files had to be iterated over a grid which resulted in 42 separate csv files. This script merges them and then merges the dataframe to the NiN data.
```{r}
# Import Sentinel NDVI Data
df.s <- list.files("P:/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/NDVI_data_Sentinel/", pattern = "*.csv", full.names=TRUE) %>%
  map_df(~fread(.))
df.s
```

##### MODIS NDVI data
The GEE code can be seen here: https://code.earthengine.google.com/efb84013701f1d5f6e1e81345f389b84
The image collection contains MODIS imagery from June, July and August 2000-2022

```{r}
# Import MODIS NDVI Data
# MODIS NDVI is scaled by 0.0001. Mean must be divided by 10000.
df.m<- read.csv("P:\\41201785_okologisk_tilstand_2022_2023\\data\\NDVI_åpenlavland\\NDVI_data_MODIS\\modis_ndvi_ts_2000_2022.csv", )
df.m
df.m$mean<-df.m$mean/10000
df.m
```

##### Landsat NDVI data
The GEE code can be seen here: https://code.earthengine.google.com/da8a9279238ef26d14be08a43788b6b7
The image collection contains Landsat imagery from June, July and August 1984-2022
To not exceed GEE memory limits, the exported files had to be iterated over a grid which resulted in 42 separate csv files. This script merges them and then merges the dataframe to the NiN data.
```{r}
# Import Landsat NDVI Data
# Set up conditional file paths
dir <- substr(getwd(), 1,2)

path <- ifelse(dir == "C:", 
               "R:/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/versjon20221231/Natur_Naturtyper_nin_norge_med_svalbard_25833/Natur_Naturtyper_NiN_norge_med_svalbard_25833.gdb",
               "/data/R/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/versjon20221231/Natur_Naturtyper_nin_norge_med_svalbard_25833/Natur_Naturtyper_NiN_norge_med_svalbard_25833.gdb")

pData <- ifelse(dir == "C:", 
                "P:/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/NDVI_data_Landsat",
                "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/NDVI_data_Landsat")

# Fread doesn't like the weird path to the server version of the P drive
# hence this horrendous work around
files=list.files(pData, pattern = "*.csv", full.names = TRUE)

df_list<-list()# initialise a list of dataframes
# read in a dataframe in each slot of the df_list
for (i in files){
  name <- gsub("-",".",i)
  name <- gsub(".csv","",name)  
  i <- paste(i,sep="")
  df_list[[i]]=assign(name,read.csv(i, header=TRUE))
}

df.l<-bind_rows(df_list, .id = "column_label")
df.l

### Data handling
